---
title: " Lumaria Life Table"
output: html_notebook
---

**Load Packages**

```{r}
library(data.table)
library(dplyr)
library(splines)
library(ggplot2)
library(scales)
```

**Download Raw Data**

```{r}
mortality_data <- data.table(read.csv("https://cdn-files.soa.org/research/srcsc/2024-srcsc-superlife-inforce-dataset.csv", skip = 3))
```

**Check Aggregate Mortality per Year**

We use the total amount of deaths / total individuals alive per year as a general proxy to identify the mortality trend over time, and in particular to check if there are any outliers.

```{r}
mortality_per_year <- c()
for (year in 2002:2023){
  deaths <- mortality_data[Year.of.Death == year,.N,]
  lives <- mortality_data[Issue.year <= year][
    (Year.of.Death >= year & is.na(Year.of.Lapse)) |
                            (Year.of.Lapse >= year & is.na(Year.of.Death) |
                               (is.na(Year.of.Death) & is.na(Year.of.Lapse))), .N]
  mortality_per_year <- c(mortality_per_year, deaths/lives)
}
plot(2002:2023,mortality_per_year)
```

**Calculate Class-Based Mortality Rates for 2015-2019**

We choose this investigation interval for recency and to exclude the outlier that is the pandemic period. I use the initial exposed to risk approach, which simply estimates q_x as the deaths during the year by those alive at the beginning of the year.

I chose to split mortality by sex.

```{r}
AM_lt <- data.table("Age" = 26:83, "Lives" = 0, "Deaths" = 0)
AF_lt <- data.table("Age" = 26:83, "Lives" = 0, "Deaths" = 0)
setkey(AM_lt, Age)
setkey(AF_lt, Age)

for (year in 2015:2019){
  year_table <- mortality_data[Issue.year <= year & Sex == "M"][
    (Year.of.Death >= year & is.na(Year.of.Lapse)) |
                            (Year.of.Lapse >= year & is.na(Year.of.Death) |
                               (is.na(Year.of.Death) & is.na(Year.of.Lapse))),][, Age := (Issue.age + year - Issue.year)][,                                            .(Lives = .N, Deaths = sum(Year.of.Death == year & !is.na(Year.of.Death))),                                                             Age]
setkey(year_table, Age)
merged_table <- year_table[AM_lt, on = "Age"] %>% replace(is.na(.), 0)
merged_table[, `:=`(Lives = Lives + i.Lives, Deaths = Deaths + i.Deaths)]
AM_lt <- merged_table[, .(Age, Lives, Deaths)]

year_table <- mortality_data[Issue.year <= year & Sex == "F"][
    (Year.of.Death >= year & is.na(Year.of.Lapse)) |
                            (Year.of.Lapse >= year & is.na(Year.of.Death) |
                               (is.na(Year.of.Death) & is.na(Year.of.Lapse))),][, Age := (Issue.age + year - Issue.year)][,                                            .(Lives = .N, Deaths = sum(Year.of.Death == year & !is.na(Year.of.Death))),                                                             Age]
setkey(year_table, Age)
merged_table <- year_table[AF_lt, on = "Age"] %>% replace(is.na(.), 0)
merged_table[, `:=`(Lives = Lives + i.Lives, Deaths = Deaths + i.Deaths)]
AF_lt <- merged_table[, .(Age, Lives, Deaths)]
}
```

Then I calculate mortality rates and check how they look. Since data is awful after age 80, I crop those ages out.

```{r}
AM_lt <- AM_lt[,q_x := Deaths/Lives]
AF_lt <- AF_lt[,q_x := Deaths/Lives]
plot(AM_lt$Age, AM_lt$q_x)
plot(AF_lt$Age, AF_lt$q_x)
AM_lt <- AM_lt[Age <= 80]
AF_lt <- AF_lt[Age <= 80]
```

They're still a bit rough, so I apply some P-Splines to graduate the raw rates. I check how nice the curves look and adjust if needed.

```{r}
model <- lm(q_x ~ bs(Age, knots = c(40, 60), degree = 3, Boundary.knots = range(Age)), data = AM_lt)
AM_lt$graduated_q_x <- predict(model, AM_lt)

model <- lm(q_x ~ bs(Age, knots = c(40, 60), degree = 3, Boundary.knots = range(Age)), data = AF_lt)
AF_lt$graduated_q_x <- predict(model, AF_lt)

plot(AM_lt$Age, AM_lt$graduated_q_x)
plot(AF_lt$Age, AF_lt$graduated_q_x)
```

Now I calculate the lapse rates. I plot them by time since the policy commenced, to see how it tracks and what would be the appropriate bins to put them in. I use a bigger window here since I don't expect the lapse rate to change much over the years. Note that only T20 policies can lapse, and all of them lapse after 20 years.

```{r}
lapse_table <- data.table("Term" = 1:19, "Lives" = 0, "Lapses" = 0)
for (year in 2001:2023){
year_table <- mortality_data[Issue.year <= year & Policy.type == "T20"][
    (Year.of.Death >= year & is.na(Year.of.Lapse)) |
                            (Year.of.Lapse >= year & is.na(Year.of.Death) |
                               (is.na(Year.of.Death) & is.na(Year.of.Lapse))),][, Term := (year - Issue.year + 1)][,.(Lives = .N,                                     Lapses = sum(Year.of.Lapse == year & !is.na(Year.of.Lapse) & is.na(Year.of.Death))), Term]
setkey(year_table, Term)
merged_table <- year_table[lapse_table, on = "Term"] %>% replace(is.na(.), 0)
merged_table[, `:=`(Lives = Lives + i.Lives, Lapses = Lapses + i.Lapses)]
lapse_table <- merged_table[, .(Term, Lives, Lapses)]
}
lapse_table <- lapse_table[,lapse_rate := Lapses/Lives]
plot(lapse_table$Term, lapse_table$lapse_rate)
```

I also need to get the policies in force split by class. I will only choose the 2023 cohort.

```{r}
policies_in_force_SPWL <- mortality_data[is.na(Year.of.Death) & is.na(Year.of.Lapse) & Policy.type == "SPWL" & Issue.year == 2023][,
                                    Age := (Issue.age + 2023 - Issue.year)
                                    ][,
                                    .(Number = .N, Sum_Insured = sum(Face.amount)),
                                    .(Policy.type, Age, Sex, Smoker.Status)]

policies_in_force_T20 <- mortality_data[is.na(Year.of.Death) & is.na(Year.of.Lapse) & Policy.type == "T20" & Issue.year == 2023][,
                                    Age := (Issue.age + 2023 - Issue.year) ][,Term := (2023 - Issue.year + 1)
                                    ][,
                                    .(Number = .N, Sum_Insured = sum(Face.amount)),
                                    .(Policy.type, Age, Term, Sex, Smoker.Status)]
```

And some plots... First, I plot raw mortality rate for age 30, 50, 70 across 2010-2023.

```{r}
mortality_over_time_30 <- c()
mortality_over_time_50 <- c()
mortality_over_time_70 <- c()
for (year in 2010:2023){
  year_table <- mortality_data[Issue.year <= year][, Age := (Issue.age + year - Issue.year)][Age == 30][
    (Year.of.Death >= year & is.na(Year.of.Lapse)) |
                            (Year.of.Lapse >= year & is.na(Year.of.Death) |
                               (is.na(Year.of.Death) & is.na(Year.of.Lapse))),][,                                            .(Lives = .N, Deaths = sum(Year.of.Death == year & !is.na(Year.of.Death))),                                                             Age]
  mortality_over_time_30 <- c(mortality_over_time_30, year_table$Deaths/year_table$Lives)
  
    year_table <- mortality_data[Issue.year <= year][, Age := (Issue.age + year - Issue.year)][Age == 50][
    (Year.of.Death >= year & is.na(Year.of.Lapse)) |
                            (Year.of.Lapse >= year & is.na(Year.of.Death) |
                               (is.na(Year.of.Death) & is.na(Year.of.Lapse))),][,                                            .(Lives = .N, Deaths = sum(Year.of.Death == year & !is.na(Year.of.Death))),                                                             Age]
  mortality_over_time_50 <- c(mortality_over_time_50, year_table$Deaths/year_table$Lives)
  
    year_table <- mortality_data[Issue.year <= year][, Age := (Issue.age + year - Issue.year)][Age == 70][
    (Year.of.Death >= year & is.na(Year.of.Lapse)) |
                            (Year.of.Lapse >= year & is.na(Year.of.Death) |
                               (is.na(Year.of.Death) & is.na(Year.of.Lapse))),][,                                            .(Lives = .N, Deaths = sum(Year.of.Death == year & !is.na(Year.of.Death))),                                                             Age]
  mortality_over_time_70 <- c(mortality_over_time_70, year_table$Deaths/year_table$Lives)
}

```

```{r}
mortality_comparison <- data.table("Year" = 2010:2023, "Age 30" = mortality_over_time_30, "Age 50" = mortality_over_time_50, "Age 70" = mortality_over_time_70)

ggplot(mortality_comparison, aes(x = Year)) + 
  geom_line(aes(y = `Age 30`, colour = "30")) +
  geom_line(aes(y = `Age 50`, colour = "50")) +
  geom_line(aes(y = `Age 70`, colour = "70")) +
    scale_colour_manual(values = c("30" = "#9370db", 
                                 "50" = "#add8e6",
                                 "70" = "red")) +
    labs(title = "Mortality Over Time, 2010-2019",
       subtitle = "Ages 30, 50 and 70",
       x = "Year",
       y = "q_x",
       colour = "Age") +
  theme_classic()
```

Then, I graph the raw and graduated mortality rates in the investigation period, across the ages analysed.

```{r}
ggplot(AF_lt, aes(x = Age)) + 
  geom_point(aes(y = q_x, colour = "Raw"), size = 1.5) + 
  geom_line(aes(y = graduated_q_x, colour = "Graduated"), size = 1.125) +
  scale_colour_manual(values = c("Raw" = "#9370db", 
                                 "Graduated" = "#add8e6")) +
  labs(title = "Assured Females, 2015-2019",
       subtitle = "Ages 26 to 80",
       x = "Age",
       y = "q_x",
       colour = "") +
  theme_classic()

ggplot(AM_lt, aes(x = Age)) + 
  geom_point(aes(y = q_x, colour = "Raw"), size = 1.5) + 
  geom_line(aes(y = graduated_q_x, colour = "Graduated"), size = 1.125) +
  scale_colour_manual(values = c("Raw" = "#9370db", 
                                 "Graduated" = "#add8e6")) +
  labs(title = "Assured Males, 2015-2019",
       subtitle = "Ages 26 to 80",
       x = "Age",
       y = "q_x",
       colour = "") +
  theme_classic()
```

Finally, I graph male and female mortality on a log scale. I do this with a file I've already saved somewhere. This will not run if you don't have the file.

```{r}
combined_genders <- read.csv("combined genders.csv")
```

```{r}
ggplot(combined_genders, aes(x = Age)) + 
  geom_line(aes(y = 100000*Males, colour = "Males")) +
  geom_line(aes(y = 100000*Females, colour = "Females")) +
    scale_colour_manual(values = c("Males" = "#add8e6", 
                                 "Females" = "#9370db")) +
    labs(title = "Male & Female Mortality Comparison",
       subtitle = "Prior to Incentives",
       x = "Age",
       y = "Deaths per 100,000",
       colour = "Sex") +
  scale_y_log10(labels = scales::comma) +
  theme_classic()
```

Now I save the data.

```{r}
write.csv(AM_lt, "Male_Life_Table.csv")
write.csv(AF_lt, "Female_Life_Table.csv")
write.csv(lapse_table, "Lapse_Table.csv")
write.csv(policies_in_force_SPWL, "Policies_in_Force_SPWL.csv")
write.csv(policies_in_force_T20, "Policies_in_Force_T20.csv")
```
